Fused regression for network inference in multiple species

### 2015-05-13 paper1.tex
intro

\title{GRN inference in multiple species}
\author{Kari Y. Lam, Zachary M. Westrick, Christian Mueller, Lionel Christiaen, Rich Bonneau}
\maketitle

\begin{abstract}
Methods for learning gene regulatory networks are essential to our understanding of cell behavior and response to external stimuli. We are interested in improving the accuracy of our networks in order to do meaningful comparisons between networks. We present two approaches using fused regression, where we exploit available evolutionary information to constrain an undetermined system, and show that both result in more accurate networks. We show that fused L2 is most useful in the case where the networks are very similar (eg different cell types or closely related species), but where data is limited. We can leverage our knowledge about the relationship between genes to promote our understanding of the respective networks. When data is more readily available, adaptive fusion is useful for identifying conservation between networks.

\end{abstract}

\section{Introduction}

>>>>

\title{Fused regression for network inference in multiple species}
\author{Kari Y. Lam, Zachary M. Westrick, Christian Mueller, Lionel Christiaen, Richard Bonneau}
\maketitle

\begin{abstract}
Learning gene regulatory networks is essential to our understanding of cell behavior and response to external stimuli. Methods for network inference have been developed and used for a wide variety of species; however, these approaches consider the problem as independent in each species even though there is known to be conservation of gene regulation even in distantly related species. We introduce a method that allows data to be pooled across species in a principled way based on gene orthology and demonstrate that it improves network inference even when the orthology is incomplete. We further refine this method with an algorithm that attempts to learn the true conserved subnetwork from a larger set of potentially conserved interactions. 
\end{abstract}

\section{Introduction}
As the volume and variety of genome scale data have exploded, the goal of accurately modeling gene regulatory networks has become more attainable. Already, there is enough data in some model organisms for the development of high quality networks; however, this is still not true of many other model systems. Luckily, the problem of inferring gene regulatory networks is not an independent problem. There has been a great deal of work showing that functional conservation exists in gene regulatory networks even across large evolutionary distance. A high quality network in one species should be a useful source when learning the network of a related species. But most existing inference algorithms do not incorporate this information.

We employ a linear model of gene expression in which the rate of expression is assumed to be proportional to a weighted sum of transcription factor expression. These weights represent the regulatory interactions we wish to learn. We can fit these weights simultaneously in multiple species of interest using time series and steady state expression data. This is similar to many existing network inference methods and produces identical answers to fitting each species separately. We introduce additional constraints on the problem that penalize differences between regulatory weights which are expected to be similar. The choice of which interactions to fuse together follows from the observation that orthologous genes are likely to have similar regulatory interactions.

A few approaches to multi-source network inference have been proposed; however, current approaches either require one-to-one orthology [aviv regev paper], which is useful for analyzing cell lines but not for most cross species comparisons, or restrict analysis to the subset of genes with orthologs in every network [ibm paper]. Our approach allows simultaneous inference of the whole gene regulatory networks of species that may have partial orthology overlap. We demonstrate performance gains in networks learned from simulated data and real data from B. subtilis and B. anthracis, where a large fraction of genes have no orthologs. Though this method was developed for multi-species network inference, it is also useful in the easier problem of learning networks in multiple cell lines.

The largest gains in performance are obtained when there is very little data in a species of interest but a large amount of data in a related species. But the information flow is two ways and our simulation results suggest that network inference improves in both species. This feature of simultaneously fitting networks is particularly useful when there is an adequate amount of data available but spread over too many species.

There exist several tools for identifying orthologous genes using sequence similarity. Although sequence similarity is a useful predictor of functional similarity, we recognize that some genes have taken on different functions and therefore may have new regulatory interactions. When this occurs, the benefits in overall performance from our method come at the cost of impaired prediction of regulators for these genes. The constraints we impose favor weight configurations in which fused interaction weights are similar to one another. Our method minimizes a cost function that strives  to simultaneously fit expression data in each species and produce networks that are consistent with these evolutionary constraints. When a gene's regulators have changed through evolution, there may be no way of achieving both these goals. In this case, the network inference procedure would be improved by relaxing or ignoring these evolutionary constraints. However, we don't know a priori which constraints to enforce and which to relax.

We approach this problem by introducing a saturating penalty function based on work in statistics to develop unbiased regularization penalties for regression. The resulting cost function is non-convex and difficult to optimize and fares less well in the underdetermined case; however, we can approximate its solution and obtain deeper insights into functional similarity than is available through orthology or the comparison of separately learned networks. When the quantity of data is sufficient to produce high quality networks, our adaptive fusion approach improves performance of network inference; this is achieved in part through the identification of presumptive orthologs with different functions. In simulated data, when we introduced faulty orthology information, our method not only correctly identified which evolutionary restraints to enforce and which to relax, but also outperformed the control case where no incorrect orthologs were used.

We present two methods to multi-species network inference using fused regression. One approach, our fused L2 method, uses an L2 fusion penalty on differences in regulatory weights with evolutionary constraints. Additionally, we developed a method, which we call adaptive fusion, using a non-convex saturating fusion penalty for both network inference and network comparison. We test the ability of fused L2 and adaptive fusion to improve network recovery on both synthetic data and Bacillus subtilis and Bacillus anthracis. We explore the circumstances under which each approach is optimal, and evaluate the robustness of adaptive fusion to incorrect orthology, simulating the biologically relevant case of neofunctionalization.
